Accelerated evolution of convective simulations

### 2018-06-13 tex/bvp_paper.tex
sections 3-4 improved.

\begin{figure}[t]
\includegraphics[width=\textwidth]{./figs/nu_v_time.png}
\caption{Here we compare the time evolution of the mean temperature and the
Nusselt number in SE (a) and AE (b) for the simulations shown in 
Fig.~\ref{fig:time_trace}. The mean value of Nu (black) is shown vs. time in both
panels, and a centered boxcar moving average with a span of 50 freefall time
units is overplotted as a thick, dashed line.  The mean value of the temperature
(blue) is overplotted. We see that the mean value of Nu rapidly approaches its
mean equilibrium value (plotted as a dashed horizontal black line), signifying
that the convective motions roughly reach a statistically stationary state quickly
compared to the temperature profile, which takes much longer to trend towards its
equilibrium value. Regardless, we see that the fluctuations of Nu about the mean
value are much smaller in the final equilibrated state than in the early state
which is not in thermal equilibrium, and that the post-AE flows capture this
proper Nu amplitude variation.
The times at which AE solves occur are denoted by arrows in (b),
and the grey boxed region is examined in more detail in Fig. \ref{fig:oscillating_plumes}.
\label{fig:nu_v_time} }
\end{figure}

@@

The Nusselt number (Nu) quantifies the efficiency of convective heat transport
and is defined as
\begin{equation}
\text{Nu} = \frac{\angles{F_{\text{conv}} + F_{\text{cond}}}}{\angles{F_{\text{cond, ref}}}}
 = \frac{\angles{wT - \mathcal{P}\partial_z T}}{\angles{- \mathcal{P} \partial_z T}},
\end{equation}
where the volume average of a quantity, $\eta$, is shown as $\angles{\eta}$.
The time evolution of Nu in AE and SE is compared to the mean temperature evolution
in Fig.~\ref{fig:nu_v_time}. The mean value of Nu evolves towards its final value
rapidly compared to the mean temperature, as shown by the thick, dashed moving
average.  As a result, if one is primarily measuring the mean value of Nu in
a convective system, a full SE run through $t_\kappa$ or the AE procedure may not
be necessary (e.g., \cite{stevens&all2010}). 
Regardless, the variations around the mean value in Nu do change significantly
over the course of thermal equilibration (Fig.~\ref{fig:nu_v_time}a). At early
times, variations in Nu about the mean are much larger than they are at later times,
and the use of AE seems to progress the fluctuations in Nu about the mean to
a state which is consistent with the equilibrated SE fluctuations.

When $S < 10^{3.67}$ in 2D and for all runs in 3D, 
the evolved system is characterized by a time-stationary value of Nu, and is thus
in a state of constant convective heat transport.
At larger $S$ in 2D, the value of Nu varies significantly over time even in the
equilibrated state,
as shown by the time traces for $S = 10^5$ (top, left) and $S = 10^7$ (bottom, left) for two
AE runs in Fig.~\ref{fig:oscillating_plumes}. We find that these systems exhibit
large Nu during
states in which temperature fluctuations travel in their natural buoyant
direction (Fig.~\ref{fig:oscillating_plumes}, Ia \& IIa, where cold elements fall and hot elements rise).
However, when wrongly-signed temperature perturbations are entrained in an upflow or downflow
with oppositely signed fluid, Nu is suppressed (Fig.~\ref{fig:oscillating_plumes}, Ib and IIb, 
where warm fluid is pulled down by the downflow
lane, and cool fluid is drawn up by the upflow lane). High-transport and 
low-transport states exhibit smaller Nu fluctuations away from the 
mean once the system reaches thermal equilibrium (Fig. \ref{fig:nu_v_time}).
The plumes in these
systems naturally oscillate horizontally over time, switching between transport being dominated
by a counterclockwise cell, as pictured in Fig.~\ref{fig:oscillating_plumes}, and
a clockwise cell. Our choice of no-slip
boundary conditions prevents the fluid from entering a full domain shearing state 
\cite{goluskin&all2014}, and the
oscillatory nature of the plumes is stable. However, thanks to our choice of periodic
boundary conditions and despite the no-slip conditions, the full system of the
upflow and downflow plumes is free to slowly migrate to the left or right over time,
and we observe this phenomenon in our solutions.
The 2D SE simulations exhibit the same horizontally
oscillatory behavior as the AE solutions for the same initial conditions. 
This time-dependent behavior of Nu is not seen strongly in our 3D solutions,
however most 3D simulations we conducted were at low $S$ compared to the runs in
which this behavior was observed in 2D.

@@

For the select 3D runs conducted in this study, the scaling of Nu, Re, and $\angles{T} - T_{\text{top}}$
reported in Fig.~\ref{fig:parameter_space_comparison}a-c is nearly identical to the
2D simulations. Errors between AE and SE solutions in 3D fall within the same range as
errors in 2D in Fig.~\ref{fig:parameter_space_comparison}d-f. AE is therefore
equally effective in both 2D and 3D, and we restrict much of our study to 2D here
in order to more thoroughly sample paramtere space.

The measurements presented in Fig.~\ref{fig:parameter_space_comparison} demonstrate
that AE can be powerfully employed in parameter space studies in which
large numbers of simulations are compared in a volume-averaged sense.  We now turn
our examination to a more direct comparison of AE and SE for 2D convection at
$S = 10^5$, the time and flux evolution of which was shown in Fig.~\ref{fig:time_trace}.
All comparisons that follow for these two runs occur over the times shaded in
green in Fig.~\ref{fig:time_trace}a\&c. Measurements are sampled every
0.1 freefall time units for 500 total freefall time units.

@@

One means of comparing two
PDFs to determine if they are drawn from the same underlying
sample distribution is through the use of a Kolmogorov-Smirnov (KS) test \cite{wall&jenkins2012}.
We calculate the KS statistic for a PDF of some value, $q$, as
\vspace{-0.5cm}
\begin{equation}
\KS{q} = \text{CDF}_{\text{AE}}(q) - \text{CDF}_{\text{SE}}(q),
\label{eqn:ks_profile}
\end{equation}
where CDF stands for cumulative distribution function, the integral of the PDF.
A traditional Kolmogorov-Smirnov statistic is just a single value,
$\KSstat{q} = |\KS{q}|_\infty =
\text{max} |\KS{q}|$, and we use both the profile KS$(q)$ and
$\KSstat{q}$ to gain insight into the likeness of two PDFs. 
We show $\KS{T}$ in Fig.~\ref{fig:temp_comparison}d, and the
CDFs used to construct it overlay the PDFs in Fig.~\ref{fig:temp_comparison}b.
Near the maxima of the temperature PDFs, $\KSstat{T} = 0.495$, 
which is very large and implies that roughly half of all
measurements in the AE case are at a lower $T$ than those in the SE case.
While this difference is significant, it is also expected from Fig.~\ref{fig:temp_comparison}a.
Fortunately, $\KS{T}$ is very small away from the maxima, 
indicating that the temperature fluctuations off of the maxima, which are the primary
drivers of convective transport, are nearly identical.

In addition to comparing the thermodynamic state achieved by the SE and AE methods,
we examine the velocities and heat transport found in the evolved states.
Shown are the PDFs of 
vertical velocity ($w$, Fig.~\ref{fig:pdf_comparison}a), horizontal velocity ($u$, Fig. \ref{fig:pdf_comparison}b),
and the nonlinear vertical convective flux ($w(T - \angles{T}_{x,y})$, Fig.~\ref{fig:pdf_comparison}c). 
Each PDF here shows a strong peak near zero due to the no-slip, impenetrable
velocity boundary conditions (Eqn.~(\ref{eqn:vel_bcs})).
The CDFs of each profile are overplotted, and corresponding KS profiles are
shown in Fig.~\ref{fig:pdf_comparison}d-f.  We report
$\KSstat{w} = 0.00615$, $\KSstat{u} = 0.0349$,
and $\KSstat{w(T - \angles{T}_{x,y})} = 0.0263$.
The difference in vertical velocity
and heat transport between AE and SE is negligible, which is unsurprising in light of
the Nu measurements of Fig.~\ref{fig:parameter_space_comparison}a\&d.
this also confirms that the large $\KSstat{T}$ in Fig.~\ref{fig:temp_comparison}d is
not of concern, and that the AE run achieves the proper convective solution.
The horizontal velocity shows some small deviation between the two simulations, and this
likely relates to the precise configuration of the convective plumes (e.g., Fig.~\ref{fig:oscillating_plumes}).
We find that the difference in $\KS{u}$, which consistently has more probability
of flows moving left (in the -$x$ direction), appears to be caused by a more prominent migration 
of the full roll system in the -$x$ direction in the AE run than in the SE run. 
This migration does not appear to affect the vertical transport appreciably.

@@

\begin{table}
\caption{We provide details regarding the computational cost of select
AE and SE runs in 2D and 3D. For each run, we quote the supercriticality
($S$), the coefficient resolution (nz, nx, ny), the number of CPUs used
to perform the calculation ($N_{\text{CPU}}$), the number of CPU-hours
used to perform the run ($t_{\text{CPU,AE/SE}}$), and the ratio of CPU-hours
used in the AE run compared to the SE run.
}
\label{table:speed}
\begin{center}
\begin{tabularx}{\textwidth}{ X X X X X X }
\hline																	
$S$	&	nz, nx$\times$ny	&	$N_{\text{CPU}}$	&	
$t_{\text{CPU,AE}}$ & $t_{\text{CPU, SE}}$ &$t_{\text{CPU,AE}}/t_{\text{CPU,SE}}$\\
\hline \hline \multicolumn{6}{c}{\vspace{-0.2cm}}\\
\multicolumn{6}{c}{\vspace{0.1cm}2D Runs} \\
\hline
$10^2$	&	64,	128	&	32    & 4.369 &  2.152    & 2.03  \\
$10^3$	&	128,256	&	64    & 20.63 &  53.25    & 0.387 \\
$10^4$	&	256,512	&	128   & 180.82&  1205.25  & 0.15  \\
$10^5$	&	512,1024&	256   & 2757.73& 24061.75 & 0.115 \\
\hline \hline \multicolumn{6}{c}{\vspace{-0.2cm}}\\
\multicolumn{6}{c}{\vspace{0.1cm}3D Runs} \\
\hline
$10^1$	&	32,	64$\times$64	& 512    &   106.15      &   61.5    & 1.73\\
$10^2$	&	64,	128$\times$128	& 512    &   111.92      &   185.35  & 0.6 \\
$10^3$	&	128,256$\times$256	& 2048   &   1426.39     &   6985.18 & 0.2 \\
$10^4$	&	256,512$\times$512	& 8192   &   23063.77    &   331,717 & 0.07\\
\hline																	
\end{tabularx}
\end{center}
\end{table}

>>>>

\begin{figure}[b]
\includegraphics[width=\textwidth]{./figs/nu_v_time.png}
\caption{The time evolution of Nu (black) and $\angles{T} - T_{\text{top}}$ (blue)
are shown for simulations at $S = 10^5$ in SE (a) and AE (b).
A moving average of Nu, using a centered boxcar with a span of 50 freefall time
units, is overplotted as a thick, dashed line. 
We see that the mean value of Nu reaches its equilibrium value quickly
compared to $\angles{T}$.
Regardless, fluctuations of Nu about the mean
value are much smaller in the thermally equilibrated state.
The times at which AE solves occur are denoted by arrows in (b),
and the grey boxed region is examined in more detail in Fig. \ref{fig:oscillating_plumes}.
\label{fig:nu_v_time} }
\end{figure}

@@

The Nusselt number (Nu) quantifies the efficiency of convective heat transport
and is defined as
\begin{equation}
\text{Nu} = \frac{\angles{F_{\text{conv}} + F_{\text{cond}}}}{\angles{F_{\text{cond, ref}}}}
 = \frac{\angles{wT - \mathcal{P}\partial_z T}}{\angles{- \mathcal{P} \partial_z T}},
\end{equation}
where the volume average of a quantity, $\eta$, is shown as $\angles{\eta}$.
The time evolution of Nu in AE and SE is compared to the mean temperature evolution
in Fig.~\ref{fig:nu_v_time}. In Fig.~\ref{fig:nu_v_time}a, we show the evolution
of the SE run at $S = 10^5$. The black trace shows the instantaneous value of
Nu, and the overplotted thick dashed line shows a moving time average of Nu. The
time average is taken using a center boxcar window whose width is 50 $t_{\text{ff}}$.
The mean value of Nu evolves towards its final value (shown in the horizontal
dashed black line) rapidly compared to the mean system temperature (shown in
blue), which approaches its evolved state (horizontal dashed blue line) slowly.
As such, it is possible to measure the statistically stationary mean value of
Nu after a few hundred freefall times, 
as has been done previously \cite{stevens&all2010}. However, the fluctuations
around the mean value are larger at early times when the system is not thermally
equilibrated. In Fig.~\ref{fig:nu_v_time}b, we show the AE traces of Nu and mean
temperature at $S = 10^5$. The times at which AE solves are completed are marked
by arrows. Nu reaches its evolved mean value slightly more rapidly than in the SE case,
but the early, large fluctuations away from the mean that we see in the SE
case are not observed after the AE solves.

When $S < 10^{3.67}$ in 2D and for all runs in 3D, 
the evolved system is characterized by a time-stationary value of Nu, and is thus
in a state of constant convective heat transport.
At larger $S$ in 2D, the value of Nu varies significantly over time even in the
equilibrated state (as seen in Fig.~\ref{fig:nu_v_time}). We examine the shaded
region of Fig.~\ref{fig:nu_v_time}b in more detail in the top left
panel of Fig.~\ref{fig:oscillating_plumes}, as well as a comparable time trace
at $S = 10^7$ (bottom, left). We find that these systems exhibit
large Nu during
states in which temperature fluctuations travel in their natural buoyant
direction (Fig.~\ref{fig:oscillating_plumes}, Ia \& IIa, where cold elements fall and hot elements rise).
However, when wrongly-signed temperature perturbations are entrained in an upflow or downflow
with oppositely signed fluid, Nu is suppressed (Fig.~\ref{fig:oscillating_plumes}, Ib and IIb, 
where warm fluid is pulled down by the downflow
lane, and cool fluid is drawn up by the upflow lane).
The plumes in these
systems naturally oscillate horizontally over time, switching between transport being dominated
by a counterclockwise cell, as pictured in Fig.~\ref{fig:oscillating_plumes}, and
a clockwise cell. Our choice of no-slip
boundary conditions prevents the fluid from entering a full domain shearing state 
\cite{goluskin&all2014}, and the
oscillatory nature of the plumes is stable. However, thanks to our choice of periodic
boundary conditions and despite the no-slip conditions, the full system of the
upflow and downflow plumes is free to slowly migrate to the left or right over time,
and we observe this phenomenon in our solutions.
The 2D SE simulations exhibit the same horizontally
oscillatory behavior as the AE solutions for the same initial conditions. 
This time-dependent behavior of Nu is not seen strongly in our 3D solutions,
however most 3D simulations we conducted were at low $S$ compared to the runs in
which this behavior was observed in 2D.

@@

For the select 3D runs conducted in this study, the scaling of Nu, Re, and $\angles{T} - T_{\text{top}}$
reported in Fig.~\ref{fig:parameter_space_comparison}a-c is nearly identical to the
2D simulations. Errors between AE and SE solutions in 3D fall within the same range as
errors in 2D in Fig.~\ref{fig:parameter_space_comparison}d-f. AE is therefore
equally effective in both 2D and 3D, and we restrict much of our study to 2D here
in order to more thoroughly sample parameter space.

The measurements presented in Fig.~\ref{fig:parameter_space_comparison} demonstrate
that AE can be powerfully employed in parameter space studies in which
large numbers of simulations are compared in a volume-averaged sense.  We now turn
our examination to a more direct comparison of AE and SE for 2D convection at
$S = 10^5$, the time, flux, and Nu evolution of which was shown in 
Figs.~\ref{fig:time_trace}, \ref{fig:nu_v_time}, \& \ref{fig:oscillating_plumes}.
All comparisons that follow for these two runs occur over the times shaded in
green in Fig.~\ref{fig:time_trace}a\&c. Measurements are sampled every
0.1 freefall time units for 500 total freefall time units.

@@

One method of comparing two
PDFs to determine if they are drawn from the same underlying
sample distribution is through the use of a Kolmogorov-Smirnov (KS) test \cite{wall&jenkins2012}.
We calculate the KS statistic for a PDF of some value, $q$, as
\vspace{-0.5cm}
\begin{equation}
\KS{q} = \text{CDF}_{\text{AE}}(q) - \text{CDF}_{\text{SE}}(q),
\label{eqn:ks_profile}
\end{equation}
where CDF stands for cumulative distribution function, the integral of the PDF.
A traditional Kolmogorov-Smirnov statistic is just a single value,
$\KSstat{q} = |\KS{q}|_\infty =
\text{max} |\KS{q}|$, and we use both the profile KS$(q)$ and
$\KSstat{q}$ to gain insight into the likeness of two PDFs. 
We show $\KS{T}$ in Fig.~\ref{fig:temp_comparison}d, and the
CDFs used to construct it overlay the PDFs in Fig.~\ref{fig:temp_comparison}b.
Near the maxima of the temperature PDFs, $\KSstat{T} = 0.495$, 
which is very large and implies that roughly half of all
measurements in the AE case are at a lower $T$ than those in the SE case.
While this difference is significant, it is also expected from Fig.~\ref{fig:temp_comparison}a.
Fortunately, $\KS{T}$ is very small away from the maxima, 
indicating that the temperature fluctuations off of the maxima, which are the primary
drivers of convective transport, are nearly identical.

In addition to adjusting the 1D thermal profile, the AE method also scales the
simulations velocities and temperature fluctuations by $\sqrt{\xi}$. 
In Fig.~\ref{fig:pdf_comparison}
we examine the velocities and heat transport found in the evolved states.
Shown are the PDFs of 
vertical velocity ($w$, Fig.~\ref{fig:pdf_comparison}a), horizontal velocity ($u$, Fig. \ref{fig:pdf_comparison}b),
and the nonlinear vertical convective flux ($w(T - \angles{T}_{x,y})$, Fig.~\ref{fig:pdf_comparison}c). 
Each PDF here shows a strong peak near zero due to the no-slip, impenetrable
velocity boundary conditions (Eqn.~(\ref{eqn:vel_bcs})).
The CDFs of each profile are overplotted, and corresponding KS profiles are
shown in Fig.~\ref{fig:pdf_comparison}d-f.  We report
$\KSstat{w} = 0.00615$, $\KSstat{u} = 0.0349$,
and $\KSstat{w(T - \angles{T}_{x,y})} = 0.0263$.
The difference in vertical velocity
and heat transport between AE and SE is negligible, which is unsurprising in light of
the Nu measurements of Fig.~\ref{fig:parameter_space_comparison}a\&d.
This also confirms that the large $\KSstat{T}$ in Fig.~\ref{fig:temp_comparison}d is
not of concern, and that the AE run achieves the proper convective solution.
The horizontal velocity shows some small deviation between the two simulations, and this
likely relates to the precise configuration of the convective plumes (e.g., Fig.~\ref{fig:oscillating_plumes}).
We find that the difference in $\KS{u}$, which consistently has more probability
of flows moving left (in the -$x$ direction), appears to be caused by a more prominent migration 
of the full roll system in the -$x$ direction in the AE run than in the SE run. 
This migration does not appear to affect the vertical transport appreciably.

@@

\begin{table}
\caption{Shown are details regarding the computational cost of select
AE and SE runs in 2D and 3D. The supercriticality
($S$), coefficient resolution (nz, nx, ny), number of CPUs used
to perform the calculation ($N_{\text{CPU}}$), number of CPU-hours
used to perform the run ($t_{\text{CPU,AE/SE}}$), and ratio of CPU-hours
used in the AE run compared to the SE run are provided.
}
\label{table:speed}
\begin{center}
\begin{tabularx}{\textwidth}{ X X X X X X }
\hline																	
$S$	&	nz, nx$\times$ny	&	$N_{\text{CPU}}$	&	
$t_{\text{CPU,AE}}$ & $t_{\text{CPU, SE}}$ &$t_{\text{CPU,AE}}/t_{\text{CPU,SE}}$\\
\hline \hline \multicolumn{6}{c}{\vspace{-0.2cm}}\\
\multicolumn{6}{c}{\vspace{0.1cm}2D Runs} \\
\hline
$10^2$	&	64,	128	&	32    & 4.369 &  2.152    & 2.03  \\
$10^3$	&	128,256	&	64    & 20.63 &  53.25    & 0.387 \\
$10^4$	&	256,512	&	128   & 180.82&  1205.25  & 0.15  \\
$10^5$	&	512,1024&	256   & 2757.73& 24061.75 & 0.115 \\
\hline \hline \multicolumn{6}{c}{\vspace{-0.2cm}}\\
\multicolumn{6}{c}{\vspace{0.1cm}3D Runs} \\
\hline
$10^1$	&	32,	64$\times$64	& 512    &   106.15      &   61.5    & 1.73\\
$10^2$	&	64,	128$\times$128	& 512    &   111.92      &   185.35  & 0.6 \\
$10^3$	&	128,256$\times$256	& 2048   &   1426.39     &   6985.18 & 0.2 \\
$10^4$	&	256,512$\times$512	& 8192   &   23063.77    &   331,717 & 0.07\\
\hline																	
\end{tabularx}
\end{center}
\end{table}
